---
title: "FMIndexR_vignette"
author: "Gabriele Oliveto"
date: " `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FMIndexR_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#>")
```

This package allows the user to efficiently build an FM index from an input FASTA file. Data structures of FM index can be also saved in a user-defined output folder. The obtained FM index can then be used to look for patterns inside a sequence, using an implementation of Backward search algorithm.

```{r setup}
library(FMIndexR)
```

Using the FM_index_from_FASTA function the FM index can easily be created. The function requires different parameters:

-   input = the input path to the FASTA file

-   output = the output path to the folder in which to save the data structures

-   save = regulates whether or not to effectively save the data structures to the output folder (default = TRUE)

The FM index is a list-like object of class "FM_index" and contains:

-   Suffix array : a data structure containing all suffixes of the input string and its associated indexes

`{FM_index_from_FASTA()}`

```{r}
sessionInfo()
```
